Constructing a universe for the setoid model

### 2020-07-04 paper.tex
small stuff

\newcommand{\metaprop}{\mathbf{Prop}}

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_P\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

@@

\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

In our presentation of the setoid model, contexts are given by setoids, that is
types together with an equivalence relation.

@@

The setoid model presented in the previous section is \emph{strict}, that is,
every equation of a CwF holds by definition in the semantics. One advantage of
strict models is that they can be turned into \emph{syntactic translations}, in
which syntactic objects of the source theory are interpreted as their
counterparts in another \emph{target} theory. In the case of the setoid model,
this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.
\footnote{%
Semantically, this translation corresponds to a model construction, in
particular a functor from the category of models of the target theory to the
category of models of what will be Setoid Type Theory.  Since the setoid
translation is structural in the context component, we can work with models in
the style of categories with families rather than contextual categories.
}

@@

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. As in the model, equality for dependent types is indexed over context
equality.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \metaprop
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

@@

We can also derive Martin-L\"of's J eliminator for this homogeneous identity
type.

@@

\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a)
\end{align*}
\end{minipage}

@@

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
%
\begin{align*}
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a setoid morphism, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:

\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}

We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

@@

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A, B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\ad{in-U~}$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\Prop_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\Prop_1$.
%
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

>>>>

@@

In addition to type formers in $\mType$, we will need the propositional versions
of $\mathbf{1}$, $\Pi$, and $\Sigma$; however these can be defined from their
$\mType$ counterparts via truncation. That is, given $P : \mProp$ and $Q : P \to
\mProp$:
%
\begin{align*}
  \Pi_p\ P\ Q & :\equiv \| \Pi\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \Sigma_p\ P\ Q & :\equiv \| \Sigma\ (\textsf{Lift} \ P)\ (\textsf{Lift} \circ Q \circ \textsf{unlift}) \| \\
  \mathbf{1}_p & :\equiv \| \mathbf{1} \|
\end{align*}

@@

\begin{align*}
  \Con &: \mType \\
  \Ty &: (\GG : \Con) \to \mType \\
  \Sub &: (\GG\ \GD : \Con) \to \mType \\
  \Tm &: (\GG : \Con) \to \Ty\ \GG \to \mType
\end{align*}

In our presentation of the setoid model, contexts are given by setoids, that is
types together with an equivalence relation. A key point of the model is that the
equivalence relation is valued in $\mProp$ and is thus definitionally proof
irrelevant.

@@

The setoid model presented in the previous section is \emph{strict}, that is,
every equation of a CwF holds by definition in the semantics. One advantage of
strict models is that they can be turned into \emph{syntactic translations}, in
which syntactic objects of the source theory are interpreted as their
counterparts in another \emph{target} theory. In the case of the setoid model,
this gives rise to a \emph{setoid translation}, where source contexts are
interpreted as target contexts together with a target type representing the
equivalence relation, and so on.\footnote{%
Semantically, this translation corresponds to a model construction, in
particular a functor from the category of models of the target theory to the
category of models of what will be Setoid Type Theory.  Since the setoid
translation is structural in the context component, we can work with models in
the style of categories with families rather than contextual categories.
}

@@

Equality type constructors for contexts and dependent types internalize the idea
that every context and type comes equipped with a setoid equivalence
relation. Note that $\mProp$ is the universe of the metatheory while $\Prop$ is
the internal one. As in the model, equality for dependent types is indexed over context
equality.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \mProp
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ \rho_{01}\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad \rho_{01} : \GG^{\sim}\ \rho_0\ \rho_1
  }
}
\]

@@

We can also derive Martin-L\"of's J eliminator for this homogeneous identity
type. The only caveat is that $\textsf{transp}$ and the J eliminator do not
compute definitionally on reflexivity.

@@

\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} :\equiv \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) :\equiv (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}

@@

$\EL$ is given by a family of setoids indexed over the universe, that is, a way
to assign to each code in the universe a carrier set and an equivalence
relation.
%
\begin{align*}
  & \El : \setoidU \to \mType \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \mProp
\end{align*}

Note that $\eqel{\_}{\_}{\_}$ is indexed over equality on the universe, because
$\El$ is a setoid morphism, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
\begin{align*}
  & \textsf{refl}_\El : (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  & \textsf{sym}_\El : \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  & \textsf{trans}_\El : \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  & \coeel : \equ{A}{B} \to \El\ A \to \El\ B \\
  & \cohel : (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}
%
We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\_\sim_\setoidU\_$, $\textsf{refl}_\setoidU$, $\El$ and $\_\vdash\_\sim_\El\_$.
The other functions are then recursively defined:
\begin{itemize}
\item $\textsf{refl}_\El$ alone,
\item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
\item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
\end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

@@

What is left now is to define the setoid equality relation on the universe, as
well as the setoid equality relation on $\El\ A$ for any $A$ in $\setoidU$.
%
Two codes $A, B$ in the universe $\setoidU$ are equal when there exists a setoid
equivalence relation on their respective sets $\El\ A$ and $\El\ B$. Intuitively,
since elements of a setoid are only ever compared to elements of the same
setoid, this should only be possible if $A$ and $B$ are codes for the same setoid,
that is, if $A \sim_{\setoidU} B$. Existence and well-formedness of such
relations is expressed via the type $\ad{in-U~}$ just defined, hence we would
expect $A \sim_{\setoidU} B$ to be defined as follows:

@@

Unfortunately this definition only manages to capture the idea, but does not
actually typecheck. In fact, $\_\sim_{\setoidU}\_$ should be $\mProp_1$-valued
relation, so $A \sim_{\setoidU} B$ should be a proposition. However, the
$\Sigma$ type shown above clearly isn't, since it quantifies over a type of
relations, which is not a proposition.
%
One possible solution is actually quite simple, and it just involves truncating
the $\Sigma$ type above to force it to be in $\mProp_1$.
%
\begin{align*}
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \mProp_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \mProp)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}
